Predicting the conformational preferences of proteins using a physics-based free energy method

### 2013-04-29 Paper/paper.tex
Made revisions

\Ken{[Guys - I recommend we get a better name than confinement.  It's totally uninformative.  In my view, better would be `confine-and-configure' or something.  
Please give it some thought.]}

@@

In some problems of protein science, you want to know the relative stability of a protein's conformation $A$ compared to it's 
conformation $B$.  We call this the \emph{difference free energy} (DFE). For example, in allosteric mechanisms, a protein often 
adopts one conformation when a ligand is bound and another conformation when no ligand is bound \cite{Elber2007}. To understand the mechanism requires knowledge 
of the relative free energies of the protein conformations.  Another example is in protein structure prediction, for example as practiced in 
the CASP competition \cite{Moult2011}.  If you are using some computational model and have predicted two putative native structures, $A$ 
and $B$, you want to compute which has the lower free energy, in order to know which is the more native-like.  In a third example, you may 
want to know whether mutating a few amino acids in a protein could cause the protein to switch from one stable conformation to another, because 
that could have important consequences for biological mechanisms and disease.  And fourth, sometimes binding a ligand induces a protein from 
one conformation to another.  A quantitative understanding of such \emph{induced-fit} situations requires knowledge of the difference free 
energy of the protein in the two states. In all these cases, it would be useful to have a computational method that can be used efficiently 
with atomically detailed physical forcefield models to compute the free energy differences between different given protein conformations.

Methods like thermodynamic integration \cite{Tironi1994},\cite{Meirovitch2007} which are successful in alchemical free energy calculation
may not be as useful for calculation of conformational free energy.
One widely explored strategy is to use molecular dynamics simulations along some putative reaction coordinate pathway from conformation $A$ 
to $B$ \cite{Elber2007,West2007,Haas2007,Jonsson1998,E2007,Dellago2002,Cheng2006}. \Ken{we need 
many refs here: cite Elber's milestoning, Jhih-Wei Chu, and the many other path-sampling approaches.]} The free energy along this reaction 
coordinate can then be determined using methods such as umbrella sampling \cite{Torrie1977}. Such approaches have several limitations.  
First, it is necessary to know a reaction pathway from $A$ to $B$. If conformations $A$ and $B$ are quite different, then it can be 
challenging to find such paths. Second, these methods 
are computationally slow. To get an accurate estimate of the total free energy difference $\Delta G = G_B - G_A$ requires accurate determinations 
of the many small free energy differences $A \rightarrow 1 \rightarrow 2 \ldots \rightarrow B$, so each microstep requires substantial amounts 
of sampling.  And third, these methods are prone to large errors, because the pathway error is a sum of many errors among the many microsteps.

\Arijit{Will it be useful to add methods like PMF or replica exchange based free energy methods ?}

\Ken{[Are all these methods below ones that don't involve paths?  Also, are there other key methods we should cite?  We need a good scholarly 
list of all the relevant works.]} 
The calculation of protein conformational free energy has been successfully attempted by a number of
groups \cite{Shell2010,Ytreberg2006,Zheng2008,Spichty2010,Strajbl2000,Park2008,Tyka2006,Cecchini2009,Ovchinnikov2013}. 
Some of these methods like reference system method \cite{Ytreberg2006}, deactivated morphing \cite{Park2008}, confinement 
method \cite{Tyka2006,Cecchini2009,Ovchinnikov2013} take an alternative strategy for computing difference free energies 
that does not require knowing a pathway from $A$ to $B$.  
\Ken{[please add the several new Karplus papers from 2012-2013.]}.

Here, we adopt the confinement method of Tyka et al \cite{Tyka2006} and Cecchini et al.\cite{Cecchini2009}. This method 
also resembles the `confine-and-release' method \cite{Mobley2007}, \cite{Mobley2012}, \cite{Mobley2006}.  
\Ken{[cite mobley's jctc 07 (ref 216 on our labsite), and mobley's recent j chem phys review, 
and mobley (ref 199) j chem phys, and others that are relevant.]}. The method is based on the thermodynamic cycle showns in 
Figure \ref{fig:method}.  To start, conformation $A$ is an ensemble.  We reduce ensemble $A$ to one particular microstate $A^\ast$ 
in that ensemble by applying restraints in an MD simulation.  In the same way, we reduce conformational ensemble $B$ to one of its 
microstates, $B^\ast$.  Then, we simply compute the free energy difference between the microstates 
as $\Delta G_{A^\ast, B^\ast} = \Delta H_{A^\ast, B^\ast}$ (which is just an enthalpy difference since there is no conformational 
entropy difference between the microstates).  $\Delta H_{A^\ast, B^\ast}$ is readily calculated using normal mode analysis. In 
this way, we can calculate the free energy difference between the two end states without needing to define a physical path or reaction 
coordinate connecting the two states.

\begin{figure}
\begin{center}
\includegraphics[width=3.5 in,height=3.5 in]{method.pdf}
\end{center}
\caption{Graphical representation of the thermodynamic cycle involving confinement method.}
\label{fig:method}
\end{figure}

\Ken{For this fig, let's make A and A* green and B and B* red or something, to distinguish them visually.  Let's also blur the ensemble more relative to the microstate, so it's clearer the distinction.]}

Previously, the confinement method has been validated on small model peptides \cite{Tyka2006,Cecchini2009}. Here, we show that the confinement 
method can also be applied successfully to larger systems, such as the series of chameleon proteins designed by Orban and co-workers
\cite{Alexander2007,He2008,Alexander2009,He2012}, which can switch between two completely different folds with only few-amino-acid changes 
in sequence.  We also show that our computed difference free energies are useful in evaluating the quality of CASP target protein predictions.  
This could be of significant value for ultimately improving the energetics in protein-structure models, not just the structures.  And, 
finally, we show that we can approximately decompose difference free energies into individual amino acid components.  This offers the 
opportunity for diagnosing the structural basis for difference free energies, which can be useful for interpreting biological mechanisms.

@@

First, we validated that our implementation of the confinement method produces results compatible with previous calculations reported in the 
literature.  The method has previously been applied to a 16 amino acid residue $\beta$-hairpin from protein G, known as 
BHP \cite{Cecchini2009}. We calculated the free energy difference between the native conformation (called bhp1), which has 
a two-stranded $\beta$-sheet and a three-stranded $\beta$-sheet called bhp3.  Our confinement calculation shows that bhp1 
is more stable by 1.7 kcal/mol, consistent with 4 $\mu$s equilibrium simulations showing that bhp1 is more favorable configuration 
by 1.8 kcal/mol, and in agreement with previous calculations \cite{Cecchini2009}.

Second, we looked at 6 target proteins from the CASP9 experiment.  For each target, we examined up to 5 submitted models.  We 
computed the difference free energy between the true native and the best model. Figure~\ref{fig:summary_casp} and 
Table~\ref{table:casp_control} shows that, in 5 
out of 6 cases, the confinement method assigns a lower free energy to the experimentally determined structure than to any of the 
decoys.  Other well-known discriminators  can also successfully make this recognition \cite{Sheffler2009}; 
 it is just an independent useful validation that the confinement calculations make sense.

\begin{figure}
\begin{center}
\includegraphics[width=3.5 in,height=3.2 in]{casp.pdf}
\end{center}
\caption{In 5 out of 6 CASP Target, the confinement method assigns a lower free energy to the experimentally determined native structure.} 
\label{fig:summary_casp}
\end{figure}

@@

\Ken{[ I think a better notation is $\Delta \Delta G = G_{best~decoy} - G_{native}$, rather than having the arrow.]}

@@

We also tested the confinement method predictions for difference free energies on the chameleon sequences of Orban 
et al. \cite{Alexander2007,He2008,Alexander2009,Bryan2010,He2012,Shortle2009}; these 
are instances in which two highly similar sequences fold into remarkably different structures. Orban and co-workers have designed a 
protein-G-like sequence of 56-residues that
is marginally stable in one of two possible folds. By mutating key residues in this sequence they
are able to stabilize one fold or the other (see Figure~\ref{fig:orban}). We call conformation ``$\beta$'' that fold into 
the protein-G-like that fold into a $4 \beta + \alpha$ fold, and we call conformation ``$\alpha$'' that fold into $3 \alpha$ helical folds.  One pair 
of sequences (GA88/GB88) is 88 percent identical in sequence, differing in seven positions. Another pair (GA95/GB95) is 95 percent identical, 
differing in three positions. 
%The third pair (GA98/GB98) differs in only a single tyrosine-to-alanine mutation. 
Accurately predicting the 
structural preferences of these structures presents a serious challenge for computational methods 
\cite{Alexander2007,He2008,Alexander2009,Bryan2010,He2012,Shortle2009}.

\begin{figure}
\includegraphics[width=5.0in]{orban.pdf}
\label{fig:orban}
\caption{Confinement method correctly predicts the structural preferences of four chameleon
sequences. The top part of the figure represent four sequences used in this study along with the
protein data bank id. Each sequence adopts either a three-helix bundle fold (denoted $\alpha$) or
$4 \beta + \alpha$ fold (denoted $\beta$). The relative free energies of
the two folds are reported for each sequence.}
\end{figure}

We initially approached this problem by making a model of each sequence with the same backbone
structure as its partner chameleon sequence. For example, we took the sequence of GA88 and built a
model with the same overall structure as GB88. We then used the confinement method to assess the
free energy difference between the experimentally determined structure of GA88 and the model (with
the GA88 sequence and the GB88 structure). The confinement method was able to predict the
conformational preferences correctly for all four sequences (data not shown). This is, however, not
surprising. It is well known \cite{Handl2009},\cite{Sheffler2009} that it is easy to distinguish 
computational models from native structures. To avoid this potential problem, we instead computed relative free energy of two
different computer generated models for each sequence. One model is based on the $\alpha$ structure and
the other on the $\beta$ structure (see Supporting Information for details on the modeling procedure).
This is a much more realistic test of the confinement method's ability to accurately calculate
relative free energies.

\Ken{[ It's not clear what this test is and why it's better, and why there are only 5 sequences.  Please clarify.]}
\Arijit{ Now there are 4 sequences as they have the experimentally available structure. The fifth sequence is discussed 
at the end of the paragraph.}

Figure~\ref{fig:orban} shows that the confinement method identifies the correct structure for all four sequences. One hypothesis for fold switching of
chameleon sequences is that the structural transitions require states with diminished stability
\cite{Bryan2010}. It is believed that if the free energy of the native state and the alternative state are
within a range of 5 kcal/mol, then it is possible for the native state to be destabilized relative
to the alternative state with only small changes in sequence. The stability of the native state can
decrease for a number reason ranging from chemical modification, breaking of disulphide bonds or
mutations---as is the case here. The calculated free energy differences range from around 3.5 to 5.0
kcal/mol, which is consistent with the above hypothesis \cite{He2008,Alexander2009,Bryan2010}. 
In a more recent study\cite{He2012}, the amino acid residue at position 45 (Tyr for $\beta$ and Leu for $\alpha$) was
found to be important for switching between  $\alpha$ and $\beta$ conformation. This inspired us to 
introduce another mutation at 45 position. For our future reference this Tyr-45-Ala mutation of GB95 sequence will
be known as GB98. Interestingly, with this mutation, the equilibrium shifts to $\alpha$ conformation as it gains 3.84 Kcal/Mol 
stability compare to $\beta$ conformation and indicate the importance of Tyr-45 residue. It will be interesting to 
know the actutal experimental structure which can validate our result.

\subsection{Per-residue free energy calculation can identify the mechanistic detail behind
conformational preferences} \label{sec:orban_per_residue}

To understand the mechanism behind the conformational preferences of the different protein
sequences, we decomposed the calculated free energy into per-residue contributions in
an approximate way \cite{Tyka2006}. For this purpose, the confinement free energy, $\Delta G_{A,A*}$ and
$\Delta G_{B,B*}$ of each residue is calculated in the usual numerical way as described
the method section. We call this method approximate as we ignore the contribution from the normal 
mode or quasiharmonic analysis. Instead, decomp module of amber is used to calculate the internal energy 
of each residue from the final restrained trajectory.

\Ken{[Arijit -- I didn't find this methods calculation.  We should at least put in a sentence or two here summarizing it.]}  
This helps us to identify important residues that stabilize a particular
conformation. The per residue free energy, $\Delta \Delta G ((\beta) - (\alpha))$ is
shown in Figure~\ref{fig:perresidue_orban}. 
In this plot, residues colored in blue favor the $\alpha$ structure, residues in red favor the $\beta$ structure, and white residues
have no preference.

\begin{figure}
\begin{center}
\includegraphics[width=4.5in]{delta_g.pdf}
\end{center}
\caption{The $\alpha$ and $\beta$ conformation associated with GA95 and GB95 sequences are colored according to the relative per residue
free energy value. In this figure, residues colored in blue favor the $\alpha$ structure, residues in red favor the $\beta$ structure, and 
white residues have no preference. Some of the mutated residues are also labelled.}
\label{fig:perresidue_orban}
\end{figure}

Although the overall free energy difference between the two structures is small (within 3-5 Kcal/Mol), the individual residues 
can show marked preferences for being in either $\alpha$ or $\beta$
conformation. Such differences can be undestood by looking in detail at the local environemnt for
those resiudes. For example, the region around residue 7 forms a random coil in the $\alpha$
structure, whereas it is forming a beta sheet structure in the $\beta$ structure. These residues
strongly favour the locally well packed and hydrogen bonded environment found in the $\beta$ sheet.
Overall, favoring $\alpha$ or $\beta$ structure is a delicate balance, were the relative global free
energy difference is small and the contributions of different per residue tendencies balance out. It
is therefore very likely that by changing key resiude preferences the global fold preferences can be
changed.

Three residues are mutated between GA95 and GB95, at positions 20, 30, and 45. In GA95, these
residues (L20, I30 and L45) stabilize the $\alpha$ structure. In GB95, two of these residues (F30
and Y45) favor the $\beta$ structure, because they have large solvent exposed surface areas in the
$\alpha$ structure, but are more buried in the $\beta$ structure. Additionally Y45 forms a hydrogen
bond with D47 in the $\beta$ structure. On the other hand, residue A20 from GB95 still favors the
$\alpha$ structure, although not as strongly as L20 in GA95.

The $\alpha$ and $\beta$ sequences are nearly identical, so there are some common features observed for all
sequences. Roles of some important residues in stabilizing either the $\alpha$ or $\beta$
conformation and possible reasons for such conformational preferences are summerized in 
supporting information Figure S1. The experimental observations
classified the protein into two parts: Amino acids 9-51 are fully structured in both folds, whereas
residues 1-8 and 52-56 are unstructured in the $\alpha$ fold, but form a $\beta$-strand in the
$\beta$ fold. Most of the amino acid residues in the region 1-9 have negative per-residue free
energies, which means that these residues favor the $\beta$ structure.

In addition to the direct effects of the mutation, there are also indirect effects due to small
perturbations due to the mutations. For example, the L20A mutation causes a slight repacking around
residue 20. This causes large changes in the per-residue free energies of nearby residues T25 and
A26. However, these differences largely cancel and the overall effect of the L20A mutation is weaker
than either the I30F or L45Y mutations.

Such per-residue free energy decompositions allow us to understand the driving forces behind protein
folding and conformational change and may be useful for designing proteins with specific structures
and functions.

\subsection{Applcation of confinement method for structure prediction}

We applied the confinement method as a ``meta-predictor'' for structure prediction. Here, the
task is to correctly identify the most accurate models out of a set of ``decoys'' generated by
different methods during the Critical Assessment of Structure Prediction (CASP) experiment. CASP is
a blind test in which different groups apply methods to predict the 3-dimensional structure of
proteins from their sequences. Each group is allowed to submit five
possible structures, which they are supposed to rank from best to worst. We have performed two
experiments centered on CASP.

In the first experiment, we examined several targets and tried to rank-order the predictions
made by the same group (same methodology used to generate the models), in others the
structures were drawn from several different groups (different methodology used). The goal of this 
experiment is to determine whether 
the physics-based confinement method can correctly identify more native-like structures as having
low free energies.

Our second experiment was to see if the confinement method can identify structures that are missed
by other meta-prediction servers. Most successful meta-prediction servers are based on the idea of
consensus: if many different prediction methods produce similar results, then that is probably a
correct prediction. This is often a powerful heuristic, but it can miss cases where there is a very
good result that is only predicted by one method. We chose several cases where such structures were
missed and assessed if the confinement method can correctly identify these accurate models.

As is common in the CASP experiment, we assess our results in terms of Global Distance Test Total
Score (GDT-TS) \cite{Zemla2003}, which is a C$\alpha$ based measure of structural accuracy. It can
be understood roughly as the percentage of residues that are correctly positioned in the model
(range 0 to 100, higher is better). The initial models for the test was chosen from different server 
groups those have traditionally done well in past CASP events.

\subsubsection{The confinement method can correctly rank different models generated by the same methodology
and distinguish the native structure}

Here, our task is to rank different submitted models of CASP9 targets that have been generated with
the same methodology. In absence of any other reliable alternative, we compare all our free energy based 
ranking with the GDT-TS based ranking which are available in CASP website.

The first test case is a protein BVU3908 from Bacteroides vulgatus whose PDB id and CASP target 
code are 2L01 and T0559 respectively.
The best predictor group for this 69 amino acid long target was "BAKER-ROSETTASERVER". We initially checked similarity
between the models submitted by them and discarded two from the analysis on
the basis of being too similar to some of the other models. The GDT\_TS score and rmsd values as
shown in Figure~\ref{fig:T0559} indicate that model 1 was predicted correctly, whereas the order of
model 3 and 5 was wrong. The main difference between model 3 and rest of the models was the  
orientation of the first alpha helix. On the other hand, as shown
in Figure~\ref{fig:T0559} the confinement method not only can differentiate the native
structure from the submitted models, the ranking also correlates perfectly with the GDT\_TS score.

@@

\begin{figure}
\begin{center}
\includegraphics[width=3.8 in,height=3.2 in]{T0559.pdf}
\end{center}
\caption{The native and three submitted model structure, along with their GDT\_TS, RMSD and relative Free energy values
of protein BVU3908 from Bacteroides vulgatus (PDB id: 2L01 and CASP code: T0559).}
\label{fig:T0559}
\end{figure}

To further test the method we continued with the example of protein BT2368 from Bacteroides
thetaiotaomicron. The PDB id of this 74 amino acid residue protein is 2L02 and CASP target code is
T0560. We compared the free energy difference between the native structure and the two of the
five submitted models from the group "Splicer". The remaining three models were discarded as 
they were too similar to the rest of the models. As shown in Figure~\ref{fig:T0560}, we can identify 
the native state correctly and our calculated free energy based ranking matches well with the
GDT\_TS score.

>>>>

\Ken{[Guys---I recommend we get a better name than confinement.  It's totally uninformative.  In my view, better would
    be `confine-and-configure' or something.  Please give it some thought.]}

\Justin{[Ken---I agree that the name isn't great, but I don't think we should change it because: it's already in the
    literature and we didn't come up with the method. I think we should consider capitalizing it as Confinement Method
    throughout.]}

@@

In some problems of protein science, you want to know the relative stability of a protein's conformation $A$ compared to
its conformation $B$.  We call this the \emph{difference free energy}. For example, in all esoteric mechanisms, a protein
often adopts one conformation when a ligand is bound and another conformation when no ligand is bound~\cite{Elber2007}.
To understand the mechanism requires knowledge of the relative free energies of the protein conformations.  Another
example is in protein structure prediction, for example as practiced in the Critical Assessment of Structure Prediction
(CASP) competition~\cite{Moult2011}. If you are using some computational model and have predicted two putative native
structures, $A$ and $B$, you want to compute which has the lower free energy, in order to know which is the more
native-like.  In a third example, you may want to know whether mutating a few amino acids in a protein could cause the
protein to switch from one stable conformation to another, because that could have important consequences for biological
mechanisms and disease. And fourth, sometimes binding a ligand induces a protein from one conformation to another. A
quantitative understanding of such \emph{induced-fit} situations requires knowledge of the difference free energy of the
protein in the two states. In all these cases, it would be useful to have a computational method that can be used
efficiently with atomically detailed physical forcefield models to compute the free energy differences between different
given protein conformations.

\Justin{[TI is actually the basis of the confinement step. We need to be more explicit about what we mean here.] Methods
    like thermodynamic integration~\cite{Tironi1994},\cite{Meirovitch2007} which are successful in alchemical free
    energy calculation may not be as useful for calculation of conformational free energy.} One widely explored strategy
is to use molecular dynamics simulations along some putative reaction coordinate pathway from conformation $A$ to
$B$~\cite{Elber2007,West2007,Haas2007,Jonsson1998,E2007,Dellago2002,Cheng2006}. \Ken{[we need many refs here: cite
    Elber's milestoning, Jhih-Wei Chu, and the many other path-sampling approaches.]} The free energy along this
reaction coordinate can then be determined using methods such as umbrella sampling \cite{Torrie1977}\Justin{[add ref
    where this is used computing free energy differences between conformations]} or targeted MD \Justin{[refs]}. Such
approaches have several limitations. First, it is necessary to know an efficient reaction pathway from $A$ to $B$. If
conformations $A$ and $B$ are quite different, then it can be challenging to find such paths. Second, these methods are
computationally slow. To get an accurate estimate of the total free energy difference $\Delta G = G_B - G_A$ requires
accurate determinations of the many small free energy differences $A \rightarrow 1 \rightarrow 2 \ldots \rightarrow B$,
so each step requires substantial amounts of sampling.  And third, these methods are prone to large errors, because the
pathway error is a sum of many errors among the many steps.

\Ken{[Are all these methods below ones that don't involve paths?  Also, are there other key methods we should cite?  We
    need a good scholarly list of all the relevant works.]}

The calculation of protein conformational free energy has been successfully attempted by a number of groups
\cite{Shell2010,Ytreberg2006,Zheng2008,Spichty2010,Strajbl2000,Park2008,Tyka2006,Cecchini2009,Ovchinnikov2013}. Some of
these methods like the reference system method~\cite{Ytreberg2006}, deactivated morphing~\cite{Park2008}, the
confinement method~\cite{Tyka2006,Cecchini2009,Ovchinnikov2013} take an alternative strategy for computing difference
free energies that does not require knowing a pathway from $A$ to $B$.  \Ken{[please add the several new Karplus papers
    from 2012-2013.]}.

Here, we adopt the confinement method of Tyka et al~\cite{Tyka2006} and Cecchini et al.~\cite{Cecchini2009}, which is
based on the thermodynamic cycle shown in Figure~\ref{fig:method}. We start with an ensemble of related conformations,
$A$. We then restrict $A$ to a much ``tighter'' ensemble $A^\ast$ by applying position restraints in a series of MD
simulations. Similarly, a second ensemble $B$ is restricted to $B^\ast$. We then compute the free energy between
$A^\ast$ and $B^\ast$ using either normal mode analysis or the quasiharmonic method [ref]. This step takes into account
both the remaining conformational entropy in each ensemble and the remaining enthalpy difference. In this way, we can
calculate the free energy difference between the two end states without needing to define a physical path or reaction
coordinate connecting the two states. The confinement method shares some similarities with the `confine-and-release'
method for computing ligand binding affinities~\cite{Mobley2007,Mobley2012,Mobley2006} \Ken{[cite mobley's jctc 07 (ref
    216 on our labsite), and mobley's recent j chem phys review, and mobley (ref 199) j chem phys, and others that are
    relevant.]}.

\begin{figure}
    \begin{center}
        \includegraphics[width=3.5 in]{method.pdf}\label{fig:method}
    \end{center}
    \caption{Graphical representation of the thermodynamic cycle involving confinement method.}
\end{figure}

Previously, the confinement method has been validated on small model peptides~\cite{Tyka2006,Cecchini2009}. Here, we
show that the confinement method can also be applied successfully to larger systems, such as the series of chameleon
proteins designed by Orban and co-workers~\cite{Alexander2007,He2008,Alexander2009,He2012}, which can switch between two
completely different folds with only slight changes in sequence. We also show that our computed difference free energies
are useful in evaluating the quality of CASP target protein predictions.  This could be of significant value for
ultimately improving the energetics in protein-structure models, not just the structures. And, finally, we show that we
can approximately decompose difference free energies into individual amino acid components. This offers the opportunity
for diagnosing the structural basis for difference free energies, which can be useful for interpreting biological
mechanisms.

@@

First, we validated that our implementation of the confinement method produces results compatible with previous
calculations reported in the literature.  The method has previously been applied to a 16 amino acid residue
$\beta$-hairpin from protein G, known as BHP~\cite{Cecchini2009}. We calculated the free energy difference between the
native conformation---called bhp1, which has a two-stranded $\beta$-sheet---and a non-native conformation---called bhp3,
which has a three-stranded $\beta$-sheet. Our confinement calculation shows that bhp1 is more stable by 1.7 kcal/mol,
consistent with 4 $\mu$s equilibrium simulations showing that bhp1 is more favorable configuration by 1.8 kcal/mol, and
in agreement with previous calculations~\cite{Cecchini2009}.

Second, we looked at 6 target proteins from the CASP9 experiment [ref].  For each target, we examined up to 5 submitted
models.  We computed the difference free energy between the true native and the best model.
Figure~\ref{fig:summary_casp} and Table~\ref{table:casp_control} show that, in 5 out of 6 cases, the confinement method
assigns a lower free energy to the experimentally determined structure than to any of the decoys.  Other well-known
discriminators can also successfully tell native structures from computer generated models~\cite{Sheffler2009} [add
rosetta holes, rosetta ff, dfire]; it is just an independent useful validation that the confinement calculations make
sense.

\Justin{[Arijit, please add GDT of best decoy to this table.]}

@@

@@

We also tested the confinement method predictions for difference free energies on the chameleon sequences of Orban et
al.!\cite{Alexander2007,He2008,Alexander2009,Bryan2010,He2012,Shortle2009}. These are instances in which two highly
similar sequences fold into remarkably different structures. Orban and co-workers have designed a protein-G-like
sequence of 56-residues that is marginally stable in one of two possible folds. By mutating key residues in this
sequence they are able to stabilize one fold or the other (see Figure~\ref{fig:orban}). We refer to the $4\beta +
\alpha$ structure as the $\beta$ conformation, and we refer to the $3\alpha$ structure as the $\alpha$ conformation. One
pair of sequences (GA88/GB88) is 88 percent identical in sequence, differing in seven positions. Another pair
(GA95/GB95) is 95 percent identical, differing in three positions. Accurately predicting the structural preferences of
these structures presents a serious challenge for computational
methods~\cite{Alexander2007,He2008,Alexander2009,Bryan2010,He2012,Shortle2009}. \Justin{Do any of these citations try to
    do any modeling or comment about how challenging this would be?}

\begin{figure}
    \begin{center}
    \includegraphics[width=5.0in]{orban.pdf}
        \label{fig:orban}
    \end{center}
    \caption{The confinement method correctly predicts the structural preferences of four chameleon sequences. The top
        part of the figure represent four sequences used in this study along with the protein data bank identifier. The
        experimentally observed fold, computationally predicted fold, and difference free energy between the two folds
        is reported for each sequence.}
\end{figure}

We initially approached this problem by making a model of each sequence with the same backbone structure as its partner
chameleon sequence. For example, we took the sequence of GA88 and built a model with the same overall structure as GB88.
We then used the confinement method to assess the free energy difference between the experimentally determined structure
of GA88 and the model (with the GA88 sequence and the GB88 structure). The confinement method was able to predict the
conformational preferences correctly for all four sequences (data not shown). This is, however, not surprising as it is
often easy to distinguish computational models from native structures~\cite{Handl2009,Sheffler2009}. To avoid this
potential problem, we instead computed relative free energy of two different computer generated models for each
sequence. One model is based on the $\alpha$ structure and the other on the $\beta$ structure (see Supporting
Information for details on the modeling procedure). This is a much more realistic test of the confinement method's
ability to accurately calculate the difference free energy.

\Ken{[ It's not clear what this test is and why it's better, and why there are only 5 sequences.  Please clarify.]}
\Justin{[There are five pairs of sequences that can be arranged into 3 different pairings. Arijit has removed the 5th
    sequence.]}

\Arijit{ Now there are 4 sequences as they have the experimentally available structure. The fifth sequence is discussed
    at the end of the paragraph.}

\Justin{[Ken: it's not that hard to tell an experimental structure from a computer generated model. Almost all of the
    knowledge-based scoring functions can do this a substantial fraction of the time. Comparing a near-native model with
    another model is a much more challenging problem.]}

Figure~\ref{fig:orban} shows that the confinement method identifies the correct structure for all four sequences. One
hypothesis is that such structural transitions require states with diminished stability~\cite{Bryan2010}. It is believed
that if the free energy of the native state and the alternative state are within a range of around 5 kcal/mol, then it
is possible for the native state to be destabilized relative to the alternative state with only small changes in
sequence~\cite{He2008,Alexander2009,Bryan2010}. The calculated free energy differences range from around 3.5 to 5.0
kcal/mol, which is consistent with this hypothesis. In a more recent study\cite{He2012}, the amino acid residue at
position 45 (Tyr for $\beta$ and Leu for $\alpha$) was found to be important for switching between $\alpha$ and $\beta$
conformation. This inspired us to introduce another mutation at this position, Y45A, which we refer to as GA98. Our
calculations predict that this mutation shifts the equilibrium to the $\alpha$ conformation, which is now more stable
than the $\beta$ conformation by 3.8 kcal/mol. Although this result has not yet been confirmed experimentally, it is
consistent with the previously observed effect of Y45L.

\subsection{Per-residue free energy calculations can identify the mechanistic detail behind conformational preferences}
\label{sec:orban_per_residue}

To better understand the mechanism behind these chameleon proteins, we decomposed the calculated free energy into
per-residue contributions in an approximate way~\cite{Tyka2006}. We decompose each confinement step ($\Delta
G_{A,A^\ast}$, $\Delta G_{B,B^\ast}$) into per-residue contributions. We can also decompose the remaining enthalpy in
the confined ensemble into per-residue contributions. However, the method is approximate because we do not include the
residual conformational entropy from the normal mode or quasiharmonic analysis steps. Computing per-residue
contributions helps us to identify important residues that stabilize a particular conformation. The per residue free
energy, $\Delta \Delta G ((\beta) - (\alpha))$ is shown in Figure~\ref{fig:perresidue_orban}.

\begin{figure}
    \begin{center}
        \includegraphics[width=4.5in]{delta_g.pdf}
        \label{fig:perresidue_orban}
    \end{center}
    \caption{Per-residue difference free energies for the $\alpha$ and $\beta$ conformations of the GA95 and GB95
        sequences. Residues colored in blue favor the $\alpha$ structure, residues in red favor the $\beta$ structure,
        and white residues have no preference.}
\end{figure}

Although the overall free energy difference between the two structures is small (within 3--5 kcal/mol), individual
residues can show marked preferences for being in either $\alpha$ or $\beta$ conformation. Such differences can be
understood by looking in detail at the local environment for those residues. For example, the region around residue 7
forms a random coil in the $\alpha$ structure, whereas it is forming a beta sheet structure in the $\beta$ structure.
These residues strongly favour the locally well packed and hydrogen bonded environment found in the $\beta$ sheet.
Overall, favoring $\alpha$ or $\beta$ structure is a delicate balance, were the relative global free energy difference
is small and the contributions of different per residue tendencies balance out. It is therefore very likely that by
changing key residue preferences the global fold preferences can be changed.

Three residues are mutated between GA95 and GB95, at positions 20, 30, and 45. In GA95, these residues (L20, I30 and
L45) stabilize the $\alpha$ structure (compare the upper and lower panels on Figure~\ref{fig:perresidue_orban}). In
GB95, two of these residues (F30 and Y45) favor the $\beta$ structure, because they have large solvent exposed surface
areas in the $\alpha$ structure, but are more buried in the $\beta$ structure. Additionally Y45 forms a hydrogen bond
with D47 in the $\beta$ structure. On the other hand, residue A20 from GB95 still favors the $\alpha$ structure,
although less strongly than L20 in GA95.

The $\alpha$ and $\beta$ sequences are nearly identical, so there are some common features observed for all sequences.
The experimental observations classified the protein into two parts: Amino acids 9--51 are fully structured in both
folds, whereas residues 1--8 and 52--56 are unstructured in the $\alpha$ fold, but form $\beta$-strands in the $\beta$
fold. Most of the amino acid residues in the region 1--9 have negative per-residue free energies, which means that these
residues favor the $\beta$ structure. Roles of some other important residues in stabilizing either the $\alpha$ or
$\beta$ conformation are summarized in Figure S1.

In addition to the direct effects of the mutation, there are also indirect effects due to small perturbations in the
environment around the mutations. For example, the L20A mutation causes a slight repacking around residue 20. This
causes large changes in the per-residue free energies of nearby residues T25 and A26. However, the changes for these two
residues have opposite signs and nearly cancel.

These per-residue free energy decompositions provide a great deal of insight into the driving forces behind protein
folding and conformational change. We believe that such calculations may also be useful for protein design---designing
proteins with specific structures and functions.

\subsection{The confinement method is a useful tool for structure prediction}

We have tested the ability of the confinement method to act as a ``meta-predictor'' for structure prediction. Here, the
task is to correctly identify the most accurate models out of a set of ``decoys'' generated by different methods during
the CASP experiment. CASP is a blind test in which different groups apply methods to predict the 3-dimensional structure
of proteins from their sequences. Each group is allowed to submit five possible structures, which they are supposed to
rank from best to worst.

We have performed two experiments centered on CASP. In the first experiment, we tried to rank-order predictions for
several targets. For each target, the predictions were either produced by a single group---presumably using the same
method for each prediction, or were produced by several different groups---using different methods. The goal of this
experiment is to determine if the physics-based confinement method can correctly identify more native-like structures as
having lower free energies. Our second experiment was to see if the confinement method can identify structures that are
missed by other meta-prediction servers. Most successful meta-prediction servers are based on the idea of consensus: if
many different prediction methods produce similar results, then that is probably a correct prediction [ref]. This is
often a powerful heuristic, but it can miss cases where there is a very good result that is only predicted by one
method. We chose several cases where such structures were missed by the best meta-predictors in CASP and assessed if the
confinement method can correctly identify these accurate models.

As is common in the CASP experiment, we assess our results in terms of Global Distance Test Total Score (GDT-TS)
\cite{Zemla2003}, which is a C$\alpha$ based measure of structural accuracy. It can be understood roughly as the
percentage of residues that are correctly positioned in the model (range 0 to 100, higher is better). We did have enough
computer model to analyze every model, so the initial models for the test were chosen from a selection of different
server groups that have done well in past CASP events.

Overall, the confinement method performs well on our CASP tests. Figure~\ref{fig:summary_casp} shows that in almost
every case, the native structure has the lowest free energy and the best model has the next lowest free energy. The
confinement method appears to be a useful tool for ranking structure predictions.

\begin{figure}
\begin{center}
\includegraphics[width=3.5 in]{casp.pdf}
\end{center}
\caption{The confinement method is usually able to identify the native structure and the best model (the model with the
    highest GDT-TS) from a set of decoys.}
\label{fig:summary_casp}
\end{figure}

\subsubsection{The confinement method can correctly rank-order structure predictions}

First, we examined the ability of the confinement method to rank different models for a target that have been generated
by a single group using the same methodology for all predictions. We examined two targets: T0559 and T0560 (see Table
?).

\Justin{[Arijit: Let's move the protein names, organisms, pdb ids, etc to a separate table and just use the CASP ids.]}

The first test case is target T0559. The best predictor group for this 69 amino acid target was ``BAKER-ROSETTASERVER''.
To save computer time, we excluded two models that were very similar to other models that we did include. For this
target, the difference free energy computed by the confinement method can be used to accurately rank-order all of the
models and the native structure (Figure~\ref{fig:T0559}).

@@

\Justin{Ken: I don't mind the 3d plots that are in here now, but I'm not sure I would like something that looks like a
    free energy landscape.}

\begin{figure}
    \begin{center}
        \includegraphics[width=3.8 in]{T0559.pdf}
        \label{fig:T0559}
    \end{center}
    \caption{The confinement method can correctly rank the native structure and three predictions produced by a single
        method.}
\end{figure}

\Justin{I'm not sure we need a second figure of this. Move it to the SI?}

We performed a similar calculation for target T0560 with two models form the group ``Splicer''. The remaining three
models were discarded as they were too similar to the rest of the models. Again, we can correctly identify the native
sate and our calculated free energy based ranking matches well with GDT-TS (Figures S?).
